---
title: "Dataframes - In Class lab"
author: "Dr. Shoemaker"
date: "DATA 2401"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Working with loaded in csv files

Use the `read.csv()` function to read the data from the `gates_donations.csv` file (available on Canvas, make sure you save it in the project repository folder) into a variable called `grants`

```{r}
## if you save it in the same folder as this file this should work. 
## But you may need edit the string to reflect the actual file name: 

grants <- read_csv("gates_donations.csv")


head(grants)
```

Because you're in an RProject, you don't need to set your working directory in RStudio! The working directory is automatically set to the repository location. (You can use `getwd()` to confirm this)

-   Use the glimpse function to look at the loaded data

```{r}
glimpse(grants)
getwd()
```

-   Create a variable `org` that contains the `organization` column of the dataset

```{r}
org <- grants$organization
```

-   Confirm that the "organization" column is a vector using the `is.vector()` function. (This is a useful debugging tip if you hit errors later!)

```{r}
is.vector(org)

```

Now you can ask some interesting questions about the dataset:

-   What was the mean grant value?

```{r}
mean(grants$total_amount)
```

-   What was the dollar amount of the largest grant?

```{r}
max(grants$total_amount)
```

-   What was the dollar amount of the smallest grant?

```{r}
min(grants$total_amount)

```

-   Which organization received the largest grant?

```{r}
grants$organization[which.max(grants$total_amount)]
```

-   Which organization received the smallest grant?

```{r}
grants$organization[which.min(grants$total_amount)]
```

-   How many grants were awarded in 2010?

```{r}
sum(grants$start_year == 2010)
```

-   How much was awarded in 2010?

```{r}

sum(grants[grants$start_year == 2010, 'total_amount'])

```

# Working with built-in data sets

Load R's "USPersonalExpenditure" dataset using the `data()` function

This will produce a data frame called `USPersonalExpenditure`

```{r}
data("USPersonalExpenditure")
```

The variable `USPersonalExpenditure` is now accessible to you. Unfortunately, it's not a data frame (it's actually a matrix, you can find that out using)

Test this using the `is.data.frame()` function:

```{r}
is.data.frame(USPersonalExpenditure)
```

Luckily, you can pass the USPersonalExpenditure variable as an argument to the `data.frame()` function to convert it a data frame.

Do this, storing the result in a new variable

```{r}
x <- data.frame(USPersonalExpenditure)
```

What are the current column names of your dataframe?

```{r}
colnames(x)
```

Consider: why are they so strange? (Think about whether you could use a number like 1940 with dollar notation!)

What are the current row names of your dataframe?

```{r}
rownames(x)
```

Add a column "Category" to your data frame that contains the rownames

```{r}
x <- rownames_to_column(x, var = "Category")
```

Consider how this data isn't tidy. Use the `pivot_longer()` function from last class to make the data tidy.

```{r}
pivot_longer(x,
             cols = c(X1940, X1945, X1950, X1955, X1960 ),
             names_to = "Year",
             values_to = "Amount_dollars"
             )
```

How much money was spent on personal care in 1940?

```{r}
sum(x$Amount_dollars[x$Category == "Personal Care" & x$Year == "X1940"])
```

How much money was spent on Food and Tobacco in 1960?

```{r}

sum(x$Amount_dollars[x$Category == "Food and Tobacco" & x$Year == "X1960"])
```

What was the highest expenditure category in 1960?

```{r}
y <- x[x$Year == "X1960", ]
y$Category[which.max(y$Amount_dollars)]
```

Define a function `lowest_category` that takes in a year as a parameter, and returns the lowest spending category of that year

```{r}



```

```{r}

```

Using your function, determine the lowest spending category of each year Hint: use the `sapply()` function to apply your function to a vector of years

```{r}

```
